Classless Java: Tuning Java Interfaces

### 2015-12-08 paper/sections/Introduction.tex
spellcheck one more time
manual check

Java 8 introduced \emph{default methods}, allowing interfaces to have
method implementations. The main motivation behind the introduction of
default methods in Java 8 is \emph{interface evolution}. That is, to
allow interfaces to be extended over time, while preserving backwards
compatibility. It soon became clear that default methods could also be
used to emulate something similar to \emph{traits}~\cite{scharli03traits}. The
original notion of traits by Scharli et al. prescribes, among other
things, that: 1) a trait provides a set of methods that implement
behaviour; and 2) a trait does not specify any state variables, so the
methods provided by traits do not access state variables
directly. Java 8 interfaces follow similar principles too. Indeed, a
detailed description of how to emulate trait-oriented programming in
Java 8 can be found in the work by Bono et al.~\cite{bono14}. The Java 8
team designing default methods, was also fully aware of that secondary
use of interfaces, but it was not their objective to model traits:
``The key goal of adding default methods to Java was "interface
evolution", not "poor man's traits"''~\cite{goetz13default}. As a result
they were happy to support the secondary use of interfaces with
default methods as long as it did not make the implementation and
language more complex.

@@

\item {\em Useful, general purpose methods, require special care in
  the presence of subtyping.} Methods such as
  \emph{fluent} setters~\cite{fowler2005fluentinterface}, not only require access to the
  internal state of an object, but they also require their return types to be
  refined in subtypes.

@@

Object interfaces do not require changes to the Java runtime or compiler, 
and they also do not introduce any new syntax (apart from a
small annotation). All three features of object interfaces are
achieved by reinterpreting existing Java syntax, and are translated
into regular Java code without loss of type-safety. Since no new
syntax is introduced, it would be incorrect to call object interfaces
a language extension or syntactic sugar. So we use the term
\emph{language tuning} instead. Language tuning sits in between a
lightweight language extension and a glorified library. Language
tuning can offer many features usually implemented by a real language
extension, but because it does not modify the language syntax
pre-existing tools can work transparently on the tuned language.  To
exploit the full-benefits of language tuning, our prototype
implementation of object interfaces uses Java annotations and AST
rewriting, allowing existing Java tools (such as IDE's) to work
out-of-the-box with our implementation. As a result we could
experiment object interfaces with several interesting Java programs,
and conduct various case studies.

@@

\item {\bf ClassLess Java (CJ):} A simple formal calculus that models 
the essential features of Java 8 interfaces with default methods, and 
can be used to formally define the translation of object interfaces. 
We prove several properties of the translation\footnote{Proofs and prototype implementation are available in
  the supplementary materials.}.

@@

\item{\bf Language Tuning:} We identify the concept of language tuning 
and we describe object interfaces as an example. We also discuss 
how other existing approaches, such as the annotations in project 
Lombok~\cite{lombok},  can be viewed as language tuning.

>>>>

Java 8 introduced \emph{default methods}, allowing interfaces to have
method implementations. The main motivation behind the introduction of
default methods in Java 8 is \emph{interface evolution}. That is, to
allow interfaces to be extended over time while preserving backwards
compatibility. It soon became clear that default methods could also be
used to emulate something similar to \emph{traits}~\cite{scharli03traits}. The
original notion of traits by Scharli et al. prescribes, among other
things, that: 1) a trait provides a set of methods that implement
behaviour; and 2) a trait does not specify any state variables, so the
methods provided by traits do not access state variables
directly. Java 8 interfaces follow similar principles too. Indeed, a
detailed description of how to emulate trait-oriented programming in
Java 8 can be found in the work by Bono et al.~\cite{bono14}. The Java 8
team designing default methods, was also fully aware of that secondary
use of interfaces, but it was not their objective to model traits:
``The key goal of adding default methods to Java was "interface
evolution", not "poor man's traits"''~\cite{goetz13default}. As a result
they were happy to support the secondary use of interfaces with
default methods as long as it did not make the implementation and
language more complex.

@@

\item {\em Useful, general purpose methods require special care in
  the presence of subtyping.} Methods such as
  \emph{fluent} setters~\cite{fowler2005fluentinterface}, not only require access to the
  internal state of an object, but they also require their return types to be
  refined in subtypes.

@@

Object interfaces do not require changes to the Java runtime or compiler, 
and they also do not introduce any new syntax (apart from a
small annotation). All three features of object interfaces are
achieved by reinterpreting existing Java syntax, and are translated
into regular Java code without loss of type-safety. Since no new
syntax is introduced, it would be incorrect to call object interfaces
a language extension or syntactic sugar. So we use the term
\emph{language tuning} instead. Language tuning sits in between a
lightweight language extension and a glorified library. Language
tuning can offer many features usually implemented by a real language
extension, but because it does not modify the language syntax
pre-existing tools can work transparently on the tuned language.  To
exploit the full-benefits of language tuning, our prototype
implementation of object interfaces uses Java annotations and AST
rewriting, allowing existing Java tools (such as IDEs) to work
out-of-the-box with our implementation. As a result we could
experiment object interfaces with several interesting Java programs,
and conduct various case studies.

@@

\item {\bf Classless Java (CJ):} A simple formal calculus that models 
the essential features of Java 8 interfaces with default methods, and 
can be used to formally define the translation of object interfaces. 
We prove several properties of the translation\footnote{Proofs and prototype implementation are available in
  the supplementary materials.}.

@@

\item{\bf Language Tuning:} We identify the concept of language tuning 
and describe object interfaces as an example. We also discuss 
how other existing approaches, such as the annotations in project 
Lombok~\cite{lombok},  can be viewed as language tuning.

### 2015-12-08 paper/sections/Overview.tex
manual check

\noindent There are some annoyances here. Firstly, the sole
purpose of classes is to provide a way to instantiate
objects. Although (in this case) it takes only one line code to
provide each of those classes, this code is essentially boilerplate
code, which does not add behaviour to the system. Secondly, 
the namespace gets filled with three additional types. For example, 
both \texttt{Horse} and \texttt{HorseImpl} are needed: \texttt{Horse}
is needed because it needs to be an interface so that \texttt{Pegasus}
can use multiple inheritance; and \texttt{HorseImpl} is needed to
provide object instantiation. 
Note that, for this very simple animal system, plain Java 8 anonymous
classes can be used to avoid these problems.  We could have simply
instantiated Pegasus using:

@@

\noindent Unfortunatelly, creating a new point object is cumbersome, even
with anonymous classes:

>>>>

\noindent There are some annoyances here. Firstly, the sole
purpose of the classes is to provide a way to instantiate
objects. Although (in this case) it takes only one line code to
provide each of those classes, this code is essentially boilerplate
code, which does not add behaviour to the system. Secondly, 
the namespace gets filled with three additional types. For example, 
both \texttt{Horse} and \texttt{HorseImpl} are needed: \texttt{Horse}
is needed because it needs to be an interface so that \texttt{Pegasus}
can use multiple inheritance; and \texttt{HorseImpl} is needed to
provide object instantiation. 
Note that, for this very simple animal system, plain Java 8 anonymous
classes can be used to avoid these problems.  We could have simply
instantiated Pegasus using:

@@

\noindent Unfortunately, creating a new point object is cumbersome, even
with anonymous classes:

### 2015-12-08 paper/sections/RelatedWork.tex
spellcheck one more time
manual check

\paragraph{Traits and Java's default methods}
Simplifying the mixins approach, traits~\cite{scharli03traits} draw a strong
line between units of reuse and object factories. Traits, as units of reusable
code, contain only methods as reusable functionality, ignoring state and state
initialization. Classes, as object factories, require functionalities from
(multiple) traits. Java 8 interfaces are closely related to
traits: concrete method implementations are allowed (via the \textbf{default}
keyword) inside interfaces. The introduction of default methods opens the gate
for various flavours of multiple inheritance in Java. Traits offer an algebra
of composition operations like sum, alias and exclusion, providing explicit conflict
resolution. Former work~\cite{bono14} provides details on mimicking the trait
algebra through Java 8 interfaces.  We briefly recall the main points of their
encoding; however we propose a different representation of \textbf{exclusion}.
The first author of~\cite{bono14} agreed (via personal communication) that
our revised version for exclusion is cleaner, typesafe and more direct.

@@

%Nevertheless, we
%believe that in other languages the separation can be more practical
%than in Java, and we do not necessarally advocate that merging the two
%notions is a better approach in general.

@@

\subsection{ThisType and MyType}
%\marco{ this section is unfair. There are unsolved problems with ThisType in negative positions, namelly it breaks subtyping. The most common solution is to allow calling some methods only when the exact type is known. This demolish most advantages of interfaces.}

@@

Scala-Virtualized~\cite{Rompf2012scalavirtualized} is an extension to Scala,
which allows blending shallow and deep embedding of DSLs. It redefines
some of Scala's language constructs to method calls,
which can be overridden by DSL implementors. Thus Scala Virtualized
can also reinterpret syntax, and be seen as a form of language tuning.
However, although many Scala's language constructs are supported,
not all language constructs can be virtualized.

>>>>

\paragraph{Traits and Java's default methods}
Simplifying the mixins approach, traits~\cite{scharli03traits} draw a strong
line between units of reuse and object factories. Traits, as units of reusable
code, contain only methods as reusable functionality, ignoring state and state
initialization. Classes, as object factories, require functionality from
(multiple) traits. Java 8 interfaces are closely related to
traits: concrete method implementations are allowed (via the \textbf{default}
keyword) inside interfaces. The introduction of default methods opens the gate
for various flavours of multiple inheritance in Java. Traits offer an algebra
of composition operations like sum, alias and exclusion, providing explicit conflict
resolution. Former work~\cite{bono14} provides details on mimicking the trait
algebra through Java 8 interfaces.  We briefly recall the main points of their
encoding; however we propose a different representation of \textbf{exclusion}.
The first author of~\cite{bono14} agreed (via personal communication) that
our revised version for exclusion is cleaner, typesafe and more direct.

@@

%Nevertheless, we
%believe that in other languages the separation can be more practical
%than in Java, and we do not necessarily advocate that merging the two
%notions is a better approach in general.

@@

\subsection{ThisType and MyType}
%\marco{ this section is unfair. There are unsolved problems with ThisType in negative positions, namely it breaks subtyping. The most common solution is to allow calling some methods only when the exact type is known. This demolishes most advantages of interfaces.}

@@

Scala-Virtualized~\cite{Rompf2012scalavirtualized} is an extension to Scala,
which allows blending shallow and deep embedding of DSLs. It redefines
some of Scala's language constructs to method calls,
which can be overridden by DSL implementors. Thus Scala-Virtualized
can also reinterpret syntax, and be seen as a form of language tuning.
However, although many Scala's language constructs are supported,
not all language constructs can be virtualized.

### 2015-12-08 paper/sections/Translation.tex
manual check

\noindent That is, we can categorize all the \emph{not implemented} methods in a pattern that we know how to implement: it is eiter a field getter, a with method or a setter method.

>>>>

\noindent That is, we can categorize all the \emph{not implemented} methods in a pattern that we know how to implement: it is either a field getter, a with method or a setter method.
